---
title: "Untargeted Metabolomics Group Comparison Analysis"
output: html_document
date: "2024-09-11"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Starting with importing and packages

```{r message=FALSE, warning=FALSE}
library(readxl) #for importing
library(reshape2)# transform
library(dplyr)# for mutate
library(ggplot2)# For PCA visualization 
library(ggfortify)# For PCA
library(tidyr) #manipulations
#library(MetaboAnalystR)### for metabolomics data not using 
library(purrr) #for row manipulaton
library(broom) #for manipulation
library(RColorBrewer) #for colorblind colors 
library(plotly) #interactive plots 
library(factoextra) #clustering
library(corrplot) # correlation
library(tidyverse)
library(viridis) #more colors
#library(h2o) #ML PCA

# --- Set the path to your exported feature table here ---
# Update this to point to your local file before knitting.
# Expected format: an .xlsx feature table exported from your LC-MS
# processing software (e.g. Compound Discoverer), with columns for
# ID, m/z, RT [min], Blank, sample intensities, and QC columns.
path <- "data/your_feature_table.xlsx"
data <- read_excel(path)
```


# Simulate the data 
```{r}
# Load necessary library
library(dplyr)

# Set the number of rows
n_rows <- 1196  # From 1001 to 2196

# Generate data
set.seed(123)  # For reproducibility
data <- data.frame(
  ID = 1001:(1001 + n_rows - 1),
  m_z = runif(n_rows, 100, 300),
  RT_min = runif(n_rows, 0.5, 3.0),
  Blank = runif(n_rows, 1, 10)
)

# Generate sample measurement columns
for (i in 1:20) {
  for (j in 1:3) {
    data[[paste0(i, "_", j)]] <- runif(n_rows, 1, 10)
  }
}

# Generate QC columns
for (k in 1:14) {
  data[[paste0("QC_", k)]] <- runif(n_rows, 1, 15)
}

# View the first few rows of the DataFrame
head(data)
```


## Reshape data

```{r}
# Transforming the data from wide to long
long_data <- melt(data, id.vars = c("ID", "m/z", "RT [min]", "Blank", colnames(data)[65:78]),
                  variable.name = "Sample", 
                  value.name = "Peak_Area") 

# Split the 'Sample' column into 'Patient' and 'Group' using string manipulation
long_data <- long_data %>%
  mutate(Patient = paste0("Patient", sub("_.*", "", Sample)),   # Extract Patient number
         Group = paste0("Group", sub(".*_", "", Sample)))       # Extract Group number

#Reorder data, and getting rid of the former columns 
long_data <- long_data[, c(1, 2, 3, 21, 22, 4, 5:ncol(long_data))]
long_data <- long_data[, -c(24, 23, 21)] 

```

## Duplicate check
```{r}
#This is a function that checks for the duplicates, you use it as check_duplicates_dataframe(long_data) 
#If it returns FALSE, there are no duplicates in the data, or TRUE, there are. 
check_duplicates_dataframe <- function(df) {
  # Ensure the input is a dataframe
  if (!is.data.frame(df)) {
    stop("Input must be a dataframe.")
  }
  
  # Identify duplicate rows by comparing lengths
  unique_rows <- unique(df)
  is_duplicate <- nrow(df) != nrow(unique_rows)
  
  return(is_duplicate)
}

check_duplicates_dataframe(long_data) #No duplicates

```

## T-tests
```{r}


# Make the group_means column based on groupping of ID and Group, calculating their means 
group_means <- long_data %>%
  group_by(ID, Group) %>%
  summarize(
    count = n(),
    mean_peak_area = gsub("\\.", "", as.numeric(mean(Peak_Area, na.rm = TRUE))),
    .groups = 'drop'
  ) %>%
    select(-count) %>%
    pivot_wider(
    names_from = Group,
    values_from = mean_peak_area,
  ) %>% 
   rename_with(tolower) %>%
   mutate(across(starts_with("group"), as.numeric))
  
options(scipen = 999) # to have the values without scientifical notation 


perform_t_tests_group <- function(data) {
  # Combine the data into one vector for each group
  group1_data <- data$group1
  group2_data <- data$group2
  group3_data <- data$group3
  
  # Perform t-tests
  t_test_group1_vs_group2 <- t.test(group1_data, group2_data, paired = TRUE) 
  t_test_group1_vs_group3 <- t.test(group1_data, group3_data, paired = TRUE)
  t_test_group2_vs_group3 <- t.test(group2_data, group3_data, paired = TRUE)
  
  # Extract p-values
  p_value_group1_vs_group2 <- tidy(t_test_group1_vs_group2)$p.value
  p_value_group1_vs_group3 <- tidy(t_test_group1_vs_group3)$p.value
  p_value_group2_vs_group3 <- tidy(t_test_group2_vs_group3)$p.value
  
  # Combine p-values into a data frame
  p_values <- data.frame(
    comparison = c("group1_vs_group2", "group1_vs_group3", "group2_vs_group3"),
    p_value = c(p_value_group1_vs_group2, p_value_group1_vs_group3, p_value_group2_vs_group3)
  )
  
  # Apply Benjamini-Hochberg correction
  p_values$adjusted_p_value <- p.adjust(p_values$p_value, method = "BH")
  
  return(p_values)
}


# Apply the function to the data
t_test_results_group <- perform_t_tests_group(group_means)

# Print the results of
print(t_test_results_group)
```


## T test for each patient with the same ID from different sample 
```{r}

perform_t_tests_by_patient <- function(long_data) {
  # Initialize a data frame to store results
  results <- data.frame()

  # Loop through each patient
  for (patient in paste0("Patient", 1:20)) {
    # Filter data for the current patient
    patient_data <- long_data %>% filter(Patient == patient)
    
    # Split data by group
    group1_data <- patient_data %>% filter(Group == "Group1") %>% pull(Peak_Area)
    group2_data <- patient_data %>% filter(Group == "Group2") %>% pull(Peak_Area)
    group3_data <- patient_data %>% filter(Group == "Group3") %>% pull(Peak_Area)

    # Skip if any group has no data
    if (length(group1_data) == 0 || length(group2_data) == 0 || length(group3_data) == 0) {
      next
    }

    
    # Perform t-tests
    
      #If the  data is paired or dependent e.g., same individual measured twice, keep the argument TRUE , if not and data is independent e.g., different individuals in two groups change it to FALSE
    
    t_test_group1_vs_group2 <- t.test(group1_data, group2_data, paired = TRUE)
    t_test_group1_vs_group3 <- t.test(group1_data, group3_data, paired = TRUE)
    t_test_group2_vs_group3 <- t.test(group2_data, group3_data, paired = TRUE)

    # Store results
    results <- rbind(results, data.frame(
      Patient = patient,
      comparison = "group1_vs_group2",
      p_value = t_test_group1_vs_group2$p.value
    ))
    
    results <- rbind(results, data.frame(
      Patient = patient,
      comparison = "group1_vs_group3",
      p_value = t_test_group1_vs_group3$p.value
    ))
    
    results <- rbind(results, data.frame(
      Patient = patient,
      comparison = "group2_vs_group3",
      p_value = t_test_group2_vs_group3$p.value
    ))
  }
  
  # Apply Benjamini-Hochberg correction
  results$adjusted_p_value <- p.adjust(results$p_value, method = "BH")

  return(results)
}


results_by_patient <- perform_t_tests_by_patient(long_data) # t test performed by means of all patients1 all ID vs the others 
print(results_by_patient)



```


## T test and log2fold and -Logpvalue
```{r}


perform_t_tests_and_log2fc <- function(long_data) {
  # Initialize a data frame to store results
  results <- data.frame()

  # Loop through each unique ID
  for (id in unique(long_data$ID)) {
    # Filter data for the current ID
    id_data <- long_data %>% filter(ID == id)
    
    # Split data by group
    group1_data <- id_data %>% filter(Group == "Group1") %>% pull(Peak_Area)
    group2_data <- id_data %>% filter(Group == "Group2") %>% pull(Peak_Area)
    group3_data <- id_data %>% filter(Group == "Group3") %>% pull(Peak_Area)

    # Check if groups have enough data
    if (length(group1_data) < 2 || length(group2_data) < 2 || length(group3_data) < 2) {
      next  # Skip this ID if any group has fewer than 2 observations
    }

    # Calculate means for log2FC
    mean_group1 <- mean(group1_data)
    mean_group2 <- mean(group2_data)
    mean_group3 <- mean(group3_data)

    
    # Perform t-tests
    
      
  #If the  data is paired or dependent e.g., same individual measured twice, keep the argument TRUE , if not and data is independent e.g., different individuals in two groups change it to FALSE
    
    t_test_group1_vs_group2 <- t.test(group1_data, group2_data, paired = TRUE)
    t_test_group1_vs_group3 <- t.test(group1_data, group3_data, paired = TRUE)
    t_test_group2_vs_group3 <- t.test(group2_data, group3_data, paired = TRUE)

    # Store results
    results <- rbind(results, data.frame(
      ID = id,
      comparison = "group1_vs_group2",
      p_value = t_test_group1_vs_group2$p.value,
      log2FC = log2(mean_group1 / mean_group2),
      neg_log10_p_value = -log10(t_test_group1_vs_group2$p.value)
    ))

    results <- rbind(results, data.frame(
      ID = id,
      comparison = "group1_vs_group3",
      p_value = t_test_group1_vs_group3$p.value,
      log2FC = log2(mean_group1 / mean_group3),
      neg_log10_p_value = -log10(t_test_group1_vs_group3$p.value)
    ))

    results <- rbind(results, data.frame(
      ID = id,
      comparison = "group2_vs_group3",
      p_value = t_test_group2_vs_group3$p.value,
      log2FC = log2(mean_group2 / mean_group3),
      neg_log10_p_value = -log10(t_test_group2_vs_group3$p.value)
    ))
  }

  # Adjust p-values using Benjamini-Hochberg method
  results$adjusted_p_value <- p.adjust(results$p_value, method = "BH")
  results$adjusted_neg_log10_p_value <- -log10(results$adjusted_p_value)

  return(results)
}

t_test_and_log2fc <- perform_t_tests_and_log2fc(long_data) #apply the function 


## bind it with RT min and m/z 

selected_columns <- long_data %>%
    select(ID, `m/z`, `RT [min]`) %>%
    distinct()  # Ensure unique rows for joining

#add the ID 
t_test_and_log2fc <- t_test_and_log2fc %>% # Join by ID 
    left_join(selected_columns, by = "ID")

```


## Here I just show the colors and the pallets we can have for the plots. 
```{r}

display.brewer.all() #check the whole pallets of colorblind colors 
display.brewer.pal(n = 11, name = "Spectral") #You can change the name to Set1 for example and you will see the other palette. 
# I like the spectral one, so I will get 2 color codes red and blue from this palette. 
brewer.pal(n = 11, name = "Spectral")# Need the color code to add it into the plot. We will get  #D53E4F and #3288BD
```

## Scatter plot - Vulcano Plot based on X Log2Fold Change and Y  -Log10P-value 
```{r}

# Create a new column for color categories that matches the rules 
t_test_and_log2fc <- t_test_and_log2fc %>%
  mutate(color_category = ifelse(p_value <= 0.05 & (log2FC > 1 | log2FC < -1), "Significant", "Not Significant"))    #Make a new column named color_category 


## Interactive plot 
Scatter <- ggplot(t_test_and_log2fc, aes(x = log2FC, y = neg_log10_p_value, color = color_category)) +
  geom_point(size = 3, alpha = 0.6, aes(text = paste("Log2 FC:", log2FC, "<br>-Log10 P-Value:", neg_log10_p_value))) +  # Add tooltip text
  labs(
    title = "Scatter Plot of Log2 Fold Change vs -Log10 P-Value",
    x = "Log2 Fold Change",
    y = "-Log10 P-Value",
    color = "Category"
  ) +
  scale_color_manual(values = c("Significant" = "#D53E4F", "Not Significant" = "#3288BD")) +
  geom_hline(yintercept = 0, color = "black", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "black", linewidth = 0.5) +
  theme_minimal(base_size = 15) +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "white")
  )

# Convert ggplot to an interactive plotly plot
interactive_plot_scatter <- ggplotly(Scatter, tooltip = c("text"))  # Use 'text' for tooltips

#Visualize it # you can click on the dots to see the actual values, or see only significant or non significant values etc 
interactive_plot_scatter

```



# Some data manipulations before starting the Boxplots and Violin plots
```{r}

# Make new DF with Unique ID and Significant color 

unique_ids <- t_test_and_log2fc %>%
  filter(color_category == "Significant") %>%
  distinct(ID)  # Extract only the unique IDs

# Display the unique IDs
#unique_ids #So we know that only those samples with those ID are significant within the long_data frame 

filtered_long_data <- long_data %>%
  filter(ID %in% unique_ids$ID)

filtered_long_data$Group <- as.factor(filtered_long_data$Group) #factor for box plots 

filtered_long_data <- filtered_long_data %>%
  mutate(Group = gsub("Group(\\d+)", "Group \\1", Group))  # Replace "Group1" with "Group 1" the same for the others 


```

## Create boxplots and violin plots for the significant samples ID 
```{r}
# Remove the # from the chunk, I added it to not have all the times the plots computed and saved
# Boxplot and Violin plot for Sample ID
# ```{r}
# 
# ids <- c(1002, 1062, 1165, 1166, 1225, 1230, 1254, 1301, 1354,
#          1436, 1437, 1459, 1505, 1506, 1659, 1778, 1779,
#          2149, 2155, 2214, 2330, 2343, 2555, 2652, 2653,
#          2707)
# 
# # Loop through each ID to create boxplots and violin plots
# for (id in ids) {
#   # Boxplot
#   boxplot_code <- ggplot(filtered_long_data %>% filter(ID == id),
#                           aes(x = Group, y = Peak_Area, fill = Group)) +
#     geom_boxplot(outlier.shape = 100) +
#     scale_fill_manual(values = c("#5E4FA2", "#9E0142", "#FEE08B")) +
#     labs(title = paste("Boxplot of Sample ID", id),
#          x = "",
#          y = "Peak Area") +
#     theme_minimal() +
#     theme(
#       plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
#       axis.title.y = element_text(face = "bold"),
#       axis.text.y = element_text(size = 12, face = "bold"),
#       axis.text.x = element_text(size = 13, face = "bold"),
#       panel.grid.major = element_blank(),
#       panel.grid.minor = element_blank()
#     )
# 
#   # Convert to interactive plot
#   interactive_boxplot <- ggplotly(boxplot_code, tooltip = c("y", "x", "Group"))
# 
#   # Save boxplot as HTML
#   htmlwidgets::saveWidget(interactive_boxplot, file = paste0("boxplot_ID_", id, ".html"), selfcontained = TRUE)
# 
#   # Violin plot
#   violin_plot_code <- ggplot(filtered_long_data %>% filter(ID == id),
#                               aes(x = Group, y = Peak_Area, fill = Group)) +
#     geom_violin(trim = TRUE) +
#     scale_fill_manual(values = c("#5E4FA2", "#9E0142", "#FEE08B")) +
#     labs(title = paste("Violin Plot of Sample ID", id),
#          x = "",
#          y = "Peak Area") +
#     theme_minimal() +
#     theme(
#       plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
#       axis.title.y = element_text(face = "bold"),
#       axis.text.y = element_text(size = 12, face = "bold"),
#       axis.text.x = element_text(size = 13, face = "bold"),
#       panel.grid.major = element_blank(),
#       panel.grid.minor = element_blank()
#     )
# 
#   # Convert to interactive plot
#   interactive_violin_plot <- ggplotly(violin_plot_code, tooltip = c("y", "x", "Group"))
# 
#   # Save violin plot as HTML
#   #htmlwidgets::saveWidget(interactive_violin_plot, file = paste0("violin_plot_ID_", id, ".html"), selfcontained = TRUE) # I COMMENTED IT SO NOT FORCE R TO SAVE THEM EVERYTIME, SINCE IT IS TIME CONSUMING
# }


```

## Some extra codes to check the max and lowest values of Peak area based on patient and group 
```{r}

 filtered_long_data %>%
  filter(ID == 1166) %>%  # Filter for the specific ID, we can replace it with anything 
  group_by(Group) %>%    # Group by the Group variable
  summarise(
    max_value = max(Peak_Area, na.rm = TRUE),
    max_patient = Patient[which.max(Peak_Area)],
    min_value = min(Peak_Area, na.rm = TRUE),
    min_patient = Patient[which.min(Peak_Area)],
    .groups = 'drop'      # Drop grouping after summarising
  )

```


## Transformation for clustering 
```{r}

#or just have 3 big groups based on patient id 
wide_data_1 <- long_data %>%
  group_by(ID,Group, Patient) %>%                         # Group by ID and Group and Patient 
  summarise(Peak_Area = mean(Peak_Area, na.rm = TRUE)) %>%  
  ungroup() %>%
  pivot_wider(names_from = Group, values_from = Peak_Area) #make it wide 


#having 20 columns for each Patient
wide_data <- long_data %>%
  group_by(ID,Group, Patient, `m/z`, `RT [min]` ) %>%   # Group by ID and Group and Patient 
  summarise(Peak_Area = mean(Peak_Area, na.rm = TRUE)) %>%  
  ungroup() %>%
  pivot_wider(names_from = Patient, values_from = Peak_Area) #make it wide 



```


#Unsupervised clustering analysis with K means 
```{r}

#scale 
#scaled_data <- wide_data_1 %>%
  #select(-ID, -Patient) %>%
  #scale()

scaled_data_2 <-wide_data %>% select(-ID, -Group) %>% scale() #data with the patient spread as each Column 


num_clusters <- 5 #change it here to edit the number of clusters 

#kmeans_result <- kmeans(scaled_data, centers = num_clusters, nstart = 50)


#wide_data_1$Cluster <- as.factor(kmeans_result$cluster)

# Visualize K-means clustering results
#fviz_cluster(kmeans_result, data = scaled_data,
             #geom = "point", ellipse.type = "convex") +
  #labs(title = "K-means Clustering of Patients")




kmeans_result_2 <- kmeans(scaled_data_2, centers = num_clusters, nstart = 50)

wide_data$Cluster <- as.factor(kmeans_result_2$cluster)


# Visualize K-means clustering results
fviz_cluster(kmeans_result_2, data = scaled_data_2,
             geom = "point", ellipse.type = "convex") +
  labs(title = "K-means Clustering of Patients")


#z<-wide_data_1 %>% filter(Cluster ==1) #highest k means, when we have separate patients in the cluster 3 

zdd<-wide_data %>% filter(Cluster ==5) #ID 2284 Highest peak 

```



## Clustering without ID 2284
```{r}

data_without_2284<- filter(wide_data, ID !=2284)

wide_data_without2284<-filter(wide_data_1, ID !=2284)


scaled_data_without_2284 <-data_without_2284 %>% select(-ID, -Group, -Cluster)  %>% scale()

num_clusters_data_without_2284 <- 5 #change it here to edit the number of clusters 

kmeans_result_2 <- kmeans(scaled_data_without_2284, centers = num_clusters, nstart = 50)


scaled_data_without_2284$Cluster <- as.factor(kmeans_result_2$cluster)

# Visualize K-means clustering results
fviz_cluster(kmeans_result_2, data = scaled_data_without_2284,
             geom = "point", ellipse.type = "convex") +
  labs(title = "K-means Clustering of Patients")


```


## Data visualization for EDTA 
```{r}

idd<-2284
boxplot_2284 <- ggplot(long_data%>% filter(ID == 2284), 
                          aes(x = Group, y = Peak_Area, fill = Group)) +
    geom_boxplot(outlier.shape = 100) +
    scale_fill_manual(values = c("#5E4FA2", "#9E0142", "#FEE08B")) +
    labs(title = paste("Boxplot of Sample ID", idd),
         x = "",
         y = "Peak Area") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      axis.text.y = element_text(size = 12, face = "bold"),
      axis.text.x = element_text(size = 13, face = "bold"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
    )


interactive_boxplot_2284 <- ggplotly(boxplot_2284, tooltip = c("y", "x", "Group"))

  # Violin plot
  violin_plot_2284 <- ggplot(long_data %>% filter(ID == 2284), 
                              aes(x = Group, y = Peak_Area, fill = Group)) +
    geom_violin(trim = TRUE) +
    scale_fill_manual(values = c("#5E4FA2", "#9E0142", "#FEE08B")) +
    labs(title = paste("Violin Plot of Sample ID", idd),
         x = "",
         y = "Peak Area") +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
      axis.title.y = element_text(face = "bold"),
      axis.text.y = element_text(size = 12, face = "bold"),
      axis.text.x = element_text(size = 13, face = "bold"),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank()
    )

  # Convert to interactive plot
  interactive_violin_2284 <- ggplotly(violin_plot_2284, tooltip = c("y", "x", "Group"))


interactive_violin_2284 


 long_data %>%
  filter(ID == 2284) %>%  # Filter for the specific ID, we can replace it with anything 
  group_by(Group) %>%    # Group by the Group variable
  summarise(
    max_value = max(Peak_Area, na.rm = TRUE),
    max_patient = Patient[which.max(Peak_Area)],
    min_value = min(Peak_Area, na.rm = TRUE),
    min_patient = Patient[which.min(Peak_Area)],
    .groups = 'drop'      # Drop grouping after summarising
  )
```

## Scatter Plots based on ID, Group and 20 patients (X axe m/z and Y axe Peak Area) DO NOT UNCOMMENT IT!!!
## IT TAKES 3 H TO PROCESS EVERYTHING OR CAN TAKE EVEN MORE IF WE GOT MORE SAMPLE ID  
```{r}

# unique_ids <- unique(wide_data$ID)  # Adjust based on your actual data frame structure 
#
# Define the groups 
# groups <- c("Group1", "Group2", "Group3") 

# Function to create plots for a specific ID and group 
# create_plot_for_id_and_group <- function(id, group) { 
#   filtered_data_2 <- wide_data %>% 
#    filter(ID == id, Group == group) %>%  # Filter by both ID and Group 
#     pivot_longer(cols = starts_with("Patient"),  # Adjust this to your patient column names 
#                  names_to = "Patient", 
#                  values_to = "Peak_Area") %>% 
#     mutate(Patient = factor(Patient, levels = paste0("Patient", 1:20)))  # Reorder levels 
# 
#   # Generate distinct colors (ensuring they are more distinct) 
#   colors <- c( -->
#     brewer.pal(8, "Dark2"), 
#     brewer.pal(8, "Set1"), 
#     brewer.pal(12, "Set3")[1:4] 
#   ) 
#   extra_colors <- c("#E41A1C", "#377EB8", "#4DAF4A", "#000000", "#FFFF33", "#A65628", "#984EA3", "#999999")
#   colors <- unique(c(colors, extra_colors)) 
#   colors <- colors[1:20]  # Select the first 20 distinct colors 
# 
#   # Create ggplot 
#   p <- ggplot(filtered_data_2, aes(x = `m/z`, y = Peak_Area, fill = Patient)) + 
#     geom_point(aes(color = Patient), size = 2) +  # Scatter plot points 
#     labs(title = paste("Scatter Plot for ID:", id, "- Group:", group), 
#          x = "m/z", 
#          y = "Peak Area", 
#          color = "Patients", 
#          fill = "Patients") + 
#     theme_minimal() + -->
#     scale_fill_manual(values = colors) +  # Apply custom colors for fill 
#     scale_color_manual(values = colors) +  # Apply custom colors for points 
#     theme( -->
#       plot.title = element_text(hjust = 0.5, face = "bold"),  # Center and bold title 
#       axis.title.x = element_text(face = "bold"),  # Bold x-axis title 
#       axis.title.y = element_text(face = "bold"),  # Bold y-axis title 
#       legend.title = element_text(size = 10) 
#     ) 
# 
#   # Convert to interactive plot 
#   interactive_plot <- ggplotly(p, tooltip = c("y", "x", "Patient")) -->
# 
  # Save interactive plot as HTML 
#   htmlwidgets::saveWidget(interactive_plot, file = paste0("scatter_plot_ID_", id, "_", group, ".html"), selfcontained = TRUE) 
# 
#   return(interactive_plot)  # Return the interactive plot if needed 
# } 
# 
# # Create plots for each unique ID and each group 
# m_zplots <- list()  # Initialize a list to store plots 
# for (id in unique_ids) { 
#   for (group in groups) { 
#     m_zplots[[paste(id, group, sep = "_")]] <- create_plot_for_id_and_group(id, group) 
#   } 
# } 
# 
# # Display the plots (optional) 
# for (plot_name in names(m_zplots)) { 
#   print(m_zplots[[plot_name]])  # Display each plot for each unique ID and group 
# } 
  
```



## Scatter Plots based on ID, Group and 20 patients (X axe RT MIN and Y axe Peak Area) 
```{r}

#  unique_ids <- unique(wide_data$ID)  # Adjust based on your actual data frame structure 
# 
# # Define the groups 
#  groups <- c("Group1", "Group2", "Group3") 
# 
# # Function to create plots for a specific ID and group 
# create_plot_for_id_and_group <- function(id, group) { 
#   filtered_data_2 <- wide_data %>% 
#     filter(ID == id, Group == group) %>%  # Filter by both ID and Group 
#     pivot_longer(cols = starts_with("Patient"),  # Adjust this to your patient column names 
#                  names_to = "Patient", 
#                  values_to = "Peak_Area") %>% 
#     mutate(Patient = factor(Patient, levels = paste0("Patient", 1:20)))  # Reorder levels 
# 
#   # Generate distinct colors (ensuring they are more distinct) 
#   colors <- c(
#     brewer.pal(8, "Dark2"), 
#     brewer.pal(8, "Set1"), 
#     brewer.pal(12, "Set3")[1:4] 
#   ) 
#   extra_colors <- c("#E41A1C", "#377EB8", "#4DAF4A", "#000000", "#FFFF33", "#A65628", "#984EA3", "#999999")
#   colors <- unique(c(colors, extra_colors)) 
#   colors <- colors[1:20]  # Select the first 20 distinct colors 
# 
#   # Create ggplot 
#   p <- ggplot(filtered_data_2, aes(x = `RT [min]`, y = Peak_Area, fill = Patient)) + 
#     geom_point(aes(color = Patient), size = 2) +  # Scatter plot points 
#     labs(title = paste("Scatter Plot for ID:", id, "- Group:", group), 
#          x = "RT [min]", 
#          y = "Peak Area", 
#          color = "Patients", 
#          fill = "Patients") + 
#     theme_minimal() + 
#     scale_fill_manual(values = colors) +  # Apply custom colors for fill 
#     scale_color_manual(values = colors) +  # Apply custom colors for points 
#     theme( 
#       plot.title = element_text(hjust = 0.5, face = "bold"),  # Center and bold title 
#       axis.title.x = element_text(face = "bold"),  # Bold x-axis title 
#       axis.title.y = element_text(face = "bold"),  # Bold y-axis title 
#       legend.title = element_text(size = 10) 
#     ) 
# 
#   # Convert to interactive plot 
#   interactive_plot <- ggplotly(p, tooltip = c("y", "x", "Patient")) 
# 
#   # Save interactive plot as HTML 
#   htmlwidgets::saveWidget(interactive_plot, file = paste0("scatter_plot_ID_", id, "_", group, ".html"), selfcontained = TRUE) 
# 
#   return(interactive_plot)  # Return the interactive plot if needed 
# } 
# 
#  # Create plots for each unique ID and each group 
#  m_zplots <- list()  # Initialize a list to store plots 
#  for (id in unique_ids) { 
#    for (group in groups) { 
#      m_zplots[[paste(id, group, sep = "_")]] <- create_plot_for_id_and_group(id, group) 
#    } 
#  } 
  
```

## PCA Attempt based on onl
```{r}


# Keep the original data "data" Remove everything except the patients (so we will have 60 columns)


colnames(data) <- gsub("QC_([1-9]|1[0-4])", "PQC\\1", gsub("([0-9]+)_([0-9]+)", "Patient\\1_Group\\2", colnames(data))) #rename 


PCA_data <- data %>%
  select(matches("Patient|Group"))

PCA_transposed <- t(PCA_data) |> as.data.frame() |> rownames_to_column(var="ID")



PCA_scaled_data <- PCA_transposed %>%
  select(-ID) %>%  # Remove the ID column
  mutate(across(everything(), ~ {
    # Log transform and scale, handling NA values
    log_transformed <- log10(. + 1)  # Log transformation
    scaled_value <- scale(log_transformed, center = TRUE, scale = TRUE)  # Centering and scaling
    return(scaled_value)  # Return the scaled value
  }))

PCA_scaled_data<-as.matrix(PCA_scaled_data) #make it matrix 

#Perform PCA and extract eigenvalues and eigenvectors
pca_result <- prcomp(PCA_scaled_data)
eigenvalues <- pca_result$sdev^2
eigenvectors <- pca_result$rotation

#  Sort the eigenvalues
sorted_indices <- order(eigenvalues, decreasing = TRUE)
sorted_eigenvalues <- eigenvalues[sorted_indices]
sorted_eigenvectors <- eigenvectors[, sorted_indices]


# Assuming you want to retain 95% of the variance
cumulative_variance <- cumsum(sorted_eigenvalues) / sum(sorted_eigenvalues)
num_components <- which.max(cumulative_variance >= 0.95)

projected_data <- as.data.frame(PCA_scaled_data %*% sorted_eigenvectors[, 1:num_components])


group<-PCA_transposed$ID  #Our group is the ID variable (that became the patients )



ordered_groups <- paste0("Patient", rep(1:20, each = 3), "_Group", 1:3)

# Ensure the group variable is a factor with the correct levels
group <- factor(group, levels = ordered_groups)

# Generate a color palette with enough colors for each unique group
colors <- viridis(length(levels(group)))  # Generate a palette with unique colors
patient_colors <- setNames(colors, levels(group))  # Name the colors according to group levels

# Create the PCA plot
pca_plot <- fviz_pca_ind(pca_result,
                          geom.ind = "point",
                          col.ind = group,
                          pointshape = 21,
                          fill.ind = "white",
                          addEllipses = TRUE,
                          legend.title = "Patient_Group") +
  scale_color_manual(values = patient_colors) +
  ggtitle("Principal Component Analysis") +  # Set the title
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center title and make it bold
    axis.title.x = element_text(face = "bold"),              # Make x-axis title bold
    axis.title.y = element_text(face = "bold")               # Make y-axis title bold
  )

interactive_pca_plot <- ggplotly(pca_plot) #Interactive PCA plot 

pca_plot 
interactive_pca_plot

```


## PCA with PQC included
```{r}

PCA_data_PQC <- data %>%
  select(matches("Patient|Group|PQC[1-9]|PQC1[0-4]"))


PCA_transposed_PQC <- t(PCA_data_PQC) |> as.data.frame() |> rownames_to_column(var="ID")



PCA_scaled_data_PQC <- PCA_transposed_PQC %>%
  select(-ID) %>%  # Remove the ID column
  mutate(across(everything(), ~ {
    # Log transform and scale, handling NA values
    log_transformed <- log10(. + 1)  # Log transformation
    scaled_value <- scale(log_transformed, center = TRUE, scale = TRUE)  # Centering and scaling
    return(scaled_value)  # Return the scaled value
  }))


PCA_scaled_data_PQC<-as.matrix(PCA_scaled_data_PQC) #make it matrix 

#Perform PCA and extract eigenvalues and eigenvectors
pca_result_PQC <- prcomp(PCA_scaled_data_PQC)
eigenvalues_PQC <- pca_result_PQC$sdev^2
eigenvectors_PQC <- pca_result_PQC$rotation

#  Sort the eigenvalues
sorted_indices_PQC <- order(eigenvalues_PQC, decreasing = TRUE)
sorted_eigenvalues_PQC <- eigenvalues_PQC[sorted_indices_PQC]
sorted_eigenvectors_PQC <- eigenvectors_PQC[, sorted_indices_PQC]


# Assuming you want to retain 95% of the variance
cumulative_variance_PQC <- cumsum(sorted_eigenvalues_PQC) / sum(sorted_eigenvalues_PQC)
num_components_PQC <- which.max(cumulative_variance_PQC >= 0.95)

projected_data_PQC <- as.data.frame(PCA_scaled_data_PQC %*% sorted_eigenvectors_PQC[, 1:num_components_PQC])


group<-PCA_transposed_PQC$ID  #Our group is the ID variable (that became the patients )



ordered_groups_PQC <- c(paste0("Patient", rep(1:20, each = 3), "_Group", 1:3), paste0("PQC", 1:14))

# Ensure the group variable is a factor with the correct levels
group_PQC <- factor(group, levels = ordered_groups_PQC)

# Generate a color palette with enough colors for each unique group
colors <- viridis(length(levels(group_PQC)))  # Generate a palette with unique colors
patient_colors_PQC <- setNames(colors, levels(group_PQC))  # Name the colors according to group levels



PCs <- pca_result_PQC$x[, 1:2]  %>% as.data.frame() # Extract PC1 and PC2
colnames(PCs) <- c("PC1", "PC2")
PCs<-as.data.frame(PCs)



PCs$PC1<- PCs$PC1-23.269449 

PCs$PC2<- PCs$PC2-3.671559 # Use the median 


PCs<-as.matrix(PCs)

pca_result_PQC$x[, 1] <-as.vector(PCs[, 1])

pca_result_PQC$x[, 2] <-as.vector(PCs[, 2])


# Create the PCA plot
pca_plot_PQC <- fviz_pca_ind(pca_result_PQC,
                          geom.ind = "point",
                          col.ind = group,
                          pointshape = 21,
                          fill.ind = "white",
                          addEllipses = TRUE,
                          legend.title = "Patient_Group_PQC") +
  scale_color_manual(values = patient_colors_PQC) +
  ggtitle("Principal Component Analysis") +  # Set the title
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center title and make it bold
    axis.title.x = element_text(face = "bold"),              # Make x-axis title bold
    axis.title.y = element_text(face = "bold")               # Make y-axis title bold
  )

interactive_pca_plot_PQC <- ggplotly(pca_plot_PQC) #Interactive PCA plot 

pca_plot_PQC 
interactive_pca_plot_PQC
```
